---
layout: post
title: Working with PostGIS in R (Part 1)
published: false
excerpt: >
  An introduction to the spatial database management system PostGIS, with a focus
  on using R as an interface to PostGIS.  
category: Spatial
tags: R Spatial GIS
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(collapse=TRUE, error = TRUE, warning = FALSE)
options(knitr.table.format = 'markdown')
```

All the packges I'll use in this post:  

```{r packages}
library(dplyr)
library(RPostgreSQL) # installation instructions below
library(sp)
library(raster)
library(rgdal) # installation instructions below
library(rgeos) # installation instructions below
library(viridis) # better colour palettes
```

[PostgreSQL](https://en.wikipedia.org/wiki/PostgreSQL) is a relational database management system, similar to MySQL, and [PostGIS](https://en.wikipedia.org/wiki/PostGIS) is an extension that adds support for geographic objects. I have frequently heard about the benefits of using spatial databases rather than a collection of shapefiles and the PostgreSQL/PostGIS combo was particularly attractive since its open source. Despite this, I've avoided using these tools because I use R almost exclusively for my spatial work and it seemed like a pain to connect the two. Well, no more, I invested the time to get these tools up and running and talking to eachother on my system.  

This is the first in a series of posts on working with PostGIS in R. Here I'll focus on getting setup.  

## Working with spatial data  

### Topology operations  

#### Union and dissolve  

```{r}
rs <- dbSendQuery(con, "
  CREATE INDEX mytable_gix
    ON mytable 
    USING GIST (geom); 
")
dbClearResult(rs)
```


https://duncanjg.wordpress.com/2012/09/23/edge-to-edge-distance-using-postgis/


```{r grid}
# r <- raster(nrows=25, ncols=25, xmn=0, xmx=1, ymn=0, ymx=1, vals=NULL)
# r <- rasterToPolygons(r)
# r <- geometry(r)
# r$continuous <- rnorm(length(r))
# r$categorical <- sample(letters[1:10], length(r), replace = T)
# spplot(r, 'continuous', col.regions = viridis(256))
```